Counting evolutions: a simple foundation for thermodynamics

### 2021-02-05 Articles/2019-Entropy/Source/thermodynamics.tex
Reviewed thermodynamics paper

In a nutshell, we characterize processes in general terms, as the description of all possible ways that the system can evolve in time, its possible evolutions, under a particular set of circumstances, as one can see in figures \ref{fig_single_evolution} through \ref{fig_with_equilibria}. Counting the possible evolutions plays a fundamental role, in much the same way that counting states plays a role in statistical mechanics. In a deterministic and reversible process, there is one to one correspondence between initial and final states: the evolutions cannot merge or diverge. If the process is deterministic but not reversible, however, the same final state will be reached from different initial states, and the evolutions will merge. If the deterministic process reaches an equilibrium, there will be no more merging. We define process entropy to be the logarithm of the count of evolutions per state, which we find to have the basic properties we associate with entropy: it is linear under system composition of independent systems, can never decrease under a deterministic process, increases if the process is non-reversible and is maximized at equilibrium. This gives us a conceptually crisp physically meaningful concept that is tied to irreversibility and equilibria in a very natural way.

@@

\section{States, processes and evolutions}

The first question we want to answer is: how can we characterize processes in the most general terms? We assume we have a system and we want to study its behavior in a given set of circumstances. If the state space $X$ of the system represents all possible configurations the system can be found in, ultimately we want to characterize how the state changes in time. As depicted in figure \ref{fig_single_evolution}, we have a copy of the state space at each moment in time, and we call an \textbf{evolution} a trajectory $x(t)$ that for each moment in time returns the state of the system.

@@

Not all evolutions will be possible under all conditions. Therefore we choose to characterize a \textbf{process} simply as the set of allowed evolutions, as we can see in figure \ref{fig_process}. Namely, a process tells us what can and cannot happen in a given set of circumstances. We also assume we have a measure $\mu$ that allows us to determine the size of a set of evolutions, to ``count'' evolutions.\footnote{The precise conceptual and mathematical definition for $\mu$ is a source of a number of technical considerations which we (will) address in a separate work. While some issues, like unit definition, are important to the physics, most have to do with the proper handling of infinities or the conditions for existence and uniqueness of the measure. Overall, they would distract from the key insights.} A state $x_t$ at a time $t$ identifies a set of of compatible evolutions $A(x_t)$, those that pass through that state at that time. That is, $x(t) = x_t$. Using $\mu$ we can give a size to that set, which we can write as $\mu(A(x_t))$ or $\mu(x_t)$ for short.

@@

Note that the probability of transitioning from one state to another during a process is linked to the allowed evolutions and their count. In fact, calculating the probability of finding the state $x_{t+\Delta t}$ at $t+\Delta t$ given $x_t$ at $t$ means determining the fraction of evolutions that pass through $x_t$ that also pass through $x_{t+\Delta t}$. That is:
\begin{equation}
	P(x_{t+\Delta t} | x_t) = \frac{\mu(A(x_t) \cap A(x_{t+\Delta t}))}{\mu(A(x_t))}.
\end{equation}
Conceptually, the ability to count evolutions is equivalent to knowing all possible transition probabilities.\footnote{Technically, a single probability measure is not enough to define all these transitions at all scales, but the overall intuition remains.}

Having answered the question of how we characterize processes, we can categorize them. For example, we say a process is \textbf{deterministic} if the state at one time is enough to determine the state at all future times. We can see an example of such a process in figure \ref{fig_determinism}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all future states: evolutions can merge but cannot diverge. The count of evolutions per state over time can never decrease: $
\mu(x(t)) \leq \mu(x(t + \Delta t))$.

Deterministic processes can be characterized by a law of evolution. That is, we can write $x(t+\Delta t) = f(x(t), t)$ given a suitable function $f$ which may, in general, be time dependent. We can do this because $P(x(t+\Delta t) | x(t)) = 0$ for all but one future state. The law of evolution identifies that state. That is, $P(x(t+\Delta t) | x(t)) = 1$ if and only if $x(t+\Delta t) = f(x(t), t)$.

@@

Conversely, we say a process is \textbf{reversible} if the state at one time is enough to reconstruct the state at all past times.\footnote{Note that in some literature the ability to reconstruct the past is called retrodictability and does not coincide with thermodynamic reversibility (the ability to undo a change). This distinction appears in the context of continuous quantities and it is precisely one of those technical considerations we (will) address in a more technical work. At this point we want to make it clear that our notion of reversibility will indeed correctly recover thermodynamic reversibility. It will do so without having to assume the existence of a reverse process or a symmetry under time-reversal.} We can see an example of such a process in figure \ref{fig_reversibility}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all past states: evolutions can diverge but cannot merge. The count of evolutions per state over time can never increase: $
\mu(x(t)) \geq \mu(x(t + \Delta t))$.

@@

A process that is both deterministic and reversible, like the one in figure \ref{fig_detrev}, will never contain evolutions that split or merge. The count of evolutions per state over time will be a constant: $\mu(x(t)) = \mu(x(t + \Delta t))$.

@@

To finish, consider a \textbf{deterministic processes with equilibria}, like the one in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{We will see later that this characterization of equilibrium is simplistic, but it will suffice for now.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

@@

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(x(t))$, that can never decrease in a deterministic process, stays the same if the process is also reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we now have two systems, with their respective state spaces $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\{x(t), y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in \ref{fig_independence}, for each possible evolution $x(t)$ of the first system and $y(t)$ of the second system, there will be a possible evolution $\{x(t), y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two systems are \textbf{independent} at a given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(x(t), y(t)) = \mu_X (x(t)) \mu_Y (y(t))$.

@@

We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under a deterministic process with equilibria. This means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems.\footnote{As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$. We will address this in the full framework.}

Under very minimal conditions, we have found a quantity that has all the right properties for entropy and that is linked with determinism and irreversibility in a plain, intuitive and physically meaningful way.

\section{System independence and state entropy}

The careful reader will have noticed that the process entropy will depend on the process, not just on the state. That is, the same description of the same system may be associated with more or fewer evolutions depending on the conditions. Entropy in physics, however, is associated to a state. Under what conditions can we associate a unique process entropy to a state?

@@

If the system is a composite there are additional considerations. Different processes may put different degrees of correlation between subsystems, therefore the same equilibria for the subsystems may correspond to different process entropy values for the composite. But if the subsystems are independent, the composite will correspond to a unique entropy value: the product.

We can then assign a unique process entropy to states of equilibria and to composite states of independent equilibria. Under these conditions we define the \textbf{state entropy} $S(x)$ which corresponds to the process entropy that the system reaches at equilibrium. But why would we want to assign a unique process entropy to states in the first place? Why would this be interesting? There are three main reasons.

The first is process combination. Suppose that we are studying a system and we change the circumstances surrounding it. It first evolves according to $A$ and at $t_0$ we switch to process $B$. If $x(t)$ is an evolution of the combined process $AB$, then for $t\leq t_0$ it must be an evolution for $A$ and for $t\geq t_0$ it must be an evolution for $B$. Moreover, the number of total evolutions must remain the same. Therefore the state $x(t_0)$, the state where we are joining the evolutions, must have a well defined entropy that is independent of the process, or not all the evolutions would match. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state.

The second reason is that, upon close inspection, measurements are never defined at an instant of time. Therefore $x(t)$, the state at an instant of time, is really a shorthand for $x[(t - dt/2, t + dt/2)]$, the state defined over a small open interval $dt$ around $t$. We can talk about a single state within that interval only if no state variables change and the system is fully characterized by those state variables alone. Therefore writing $x(t)$ implicitly assumes two things: that the system is an equilibrium of the processes that happen at a faster scale than our measurements of said system; that further description (e.g. the internal dynamics of the constituents at a finer resolution) is not needed to describe the process.

The third reason is that by state we implicitly mean a description of a system and only about that system. If there are correlations between that system and others at a time, then the description of one will tell us more than that system alone. Proper states are then defined in the case where the system is independent, where all the correlations have been removed. This is true also for correlations with the environment and correlations between subsystems. But the condition of independence is precisely the condition that maximizes entropy, making it uniquely defined.

All three insights point toward the same direction: the very notion of state seems to be intrinsically associated to the notion of equilibrium and state entropy. This would suggest that all states for all systems are, in a sense, equilibria. But what exactly is this sense?  What does it mean that a mechanical system, classical or quantum, is at equilibrium?

Even in thermodynamics, equilibrium is usually an ill-defined concept. It is typically associated with the one coming from dynamical systems (i.e. the system no longer changes), but this doesn't quite work. In fact, it is not even relativistically consistent. If we have a volume of gas in thermodynamic equilibrium, we would like to say it is in equilibrium for any observer, not just the one in its rest frame. A boosted observer, then, would see an equilibrium where the position is indeed changing. In the presence of gravitational forces, this motion would not even be linear. The notion of equilibrium as ``nothing is changing'', then, is not quite right and should be amended. But how? The previous discussion would hint toward system independence, but why should system independence imply an equilibrium?

The insight is that we can define something as a system only when we are able to take it apart from the rest and study it independently. Therefore we need, at least conceptually, a process that takes that something and makes it independent from the rest, decouples it, makes it autonomous. Our ability to define a system implicitly assumes the existence of such a ``decoupling'' process. Note that if the system is already independent, the decoupling process does nothing: in this sense independent systems are an equilibrium of a decoupling process.

This ``decoupling'' process is what maximizes entropy. If the system is truly rendered independent, then we cannot know more than the description given by the state at that particular time. We cannot know how the system came to be in that state, we cannot know anything more about the subsystems, other systems or the environment than what the state of the system strictly tell us.   When the system is independent, the states tell us the least possible information about everything else and the evolution of anything else.

The definition of a particular system, then, is conditional on the existence of particular processes: those that render the system well defined. We can talk about the motion of a ball because we are on earth at 20 Celsius. If we were on the surface of the sun, we would not be able to talk about the ball in the first place. To define a system requires us to define a boundary, which should remain the same during the process. In mathematical terms, the state space must be a symmetry (i.e. an invariant, an equilibrium) of a group (i.e. a set of transformations, a set of processes).

The full mathematical development of these concepts is left for another work, but the conceptual foundations are simple and direct. Grouping quantities into cohesive units that ``belong together'', into systems, comes with a set of practical requirements: we need a way to prepare and study them independently. This notion of ``independence equilibrium'' is what we will use as a foundation for ``thermodynamic equilibrium'', and is generalized to all systems. It is the necessary and sufficient condition to assign state entropies. In this light, entropy maximization is not a universal law of the universe, but a necessary requirement to build models. We can only study the state of objects that allow process entropy maximization. What remains is to answer one last fundamental question: is this notion of state entropy at all related to thermodynamic entropy?

\section{Thermodynamics}

The notion of equilibrium as independence is in general not sufficient to recover thermodynamic equilibrium. So, how are they related? When does the second arise?

We saw how the notion of state as independence equilibrium came naturally when considering process composition of a single system. When the process changes, we need to reconcile the count of evolution at the junction. We now want to consider the scenario of process composition of composite systems. The idea is that we want to describe the situation in which two systems come together, evolve as a composite for some time, and then proceed to evolve separately. In this case, at each juncture we have to reconcile the evolutions of both the whole and each of the parts. We need to have an independence equilibrium at both levels. What we'll find is that thermodynamics studies this case. But let's proceed in order.

If our goal is to characterize part-whole relationships, a crucial question is: how are the state variables of the whole related to the state variables of the parts? We define a \textbf{thermodynamic system} one that follows these rules: 
\begin{itemize}
	\item states (i.e. independence equilibria) are identified by a set of extensive quantities (i.e. additive under system composition)\footnote{It may be possible that the existence of extensive quantities can be derived from the composability of thermodynamic equilibria. Note, in fact, that we typically construct measurement scales of quantities (such as energy, mass, volume, distance, ...) by preparing unit samples and then adding or dividing them. Assuming the existence of thermodynamic equilibria may enable us similar constructions without extra assumptions. (XXX: any work related to this?) Nevertheless, for this work, we consider the existence of extensive quantities an extra assumption.
	}
	\item one such quantity is called internal energy
	\item internal energy will be conserved under any deterministic process for that system.
\end{itemize}
We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities.

For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities. This is simply the equation of state in entropic form, which is the starting point of Gibbs formulation of thermodynamics. [CITE] We can therefore define the relationships
\begin{align}
\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ is dimensionless.

As we need to characterize states of both whole and parts, we need to characterize the part-whole relationship of the equilibrium. We say that a thermodynamic system $A$ in \textbf{thermodynamic equilibrium} has the following properties:
\begin{itemize}
	\item the system $A$ is in an independence equilbrium
	\item all parts of $A$ are in thermodynamic equilibrium
	\item if another composite system $B$ is in thermodynamic equilibrium and $A$ and $B$ share a part, then the composite of $A$ and $B$ is in thermodynamic equilibrium.
\end{itemize}
We say two or more systems are in thermodynamic equilibrium with each other if their composite is in thermodynamic equilibrium.\footnote{To understand the mathematical structure a bit better, the set of all systems $S$ is a Boolean algebra. Element $0$ represents ``nothing'' while element $1$ represents ``everything''. The join $\OR$ operation is system composition while the meet $\AND$ returns the smallest system contained by the two. Therefore $A \OR 1 = 1$: any system composed with everything gives us everything. As a lattice, the partial order $\leq$ is system inclusion. A thermodynamic equilibrium essentially identifies a topmost system, with all the subsystem and only the subsystems as part of the same equilibrium. That is, for every thermodynamic equilibrium there exists a system $A$ such that every $B \leq A$ is in the same equilibrium while every $C \nleq A$ is not in the same equilibrium.} A \textbf{thermodynamic process} is a process with equilibria where the final state is a thermodynamic equilibrium. We stress that we are not saying that two systems, in general, always allow for thermodynamic equilibria. However, we contend that these are the type of processes we can mix and match to create composite processes of composite systems. What we want to explore is what happens if we had such systems and processes, at least to a first approximation.

To get a feel for the definition, let us rederive the fact that thermodynamic equilibrium is transitive, also known as the zeroth law. Suppose system $A$ is in thermodynamic equilibrium with system $B$ and system $B$ is in thermodynamic equilibrium with system $C$. Then the composites $A \times B$ and $B \times C$ are respectively in thermodynamic equilibrium. Since they share a part, $B$, this means that $A \times B \times C$ is in thermodynamic equilibrium. But $A \times C$ is a part of $A \times B \times C$, so it is in thermodynamic equilibrium. Which means $A$ is in thermodynamic equilibrium with $C$.

The characterization of thermodynamic equilibrium we give is manifestly a part-whole relationship. It extends the notion of independence equilibrium to a composite system so that the equilibrium of the whole and the equilibrium of its parts are essentially the same equilibrium. We find that clarifying this kind of part-whole relationships makes for a good conceptual starting point not only for thermodynamics, but also for the other branches of physics. This should not come as a surprise as reductionism (i.e. explaining the whole through its parts) plays an important role in science.

%Under these assumptions/definitions, we can recover the laws of thermodynamics. Let $A$ and $B$ be two thermodynamic system in thermodynamic equilibrium and let $AB$ be the composite. We have:
%\begin{equation}
%\begin{aligned}
%U_{AB} &= U_A + U_B \\
%S_{AB}(U_{AB}) &= S_A(U_A) + S_B(U_B) \\
%&= \sup \{S_A(x) + S_B(y) \, | \, x+y= U_{AB}\} \\
%\frac{1}{kT_A} = \beta_A = \frac{\partial S_A}{\partial U_A} &= \frac{\partial S_B}{\partial U_B} = \beta_B = \frac{1}{kT_B}
%\end{aligned}
%\end{equation}
%which follows the standard textbooks derivations and assumes the new equilibrium is not on a boundary point. In this case, equilibrium is achieved if and only if $T_A$ and $T_B$ are equal. Since equilibria are unique, this will also tell us that all equations of states have to be concave in $U$. If $B$ is in thermal equilibrium with another system $C$ we also have $T_B$ is equal to $T_C$ and therefore $A$ and $C$ are also in thermal equilibrium. This recovers the zeroth law.
%
%There is, however, a problem. We assumed the equilibrium is not a boundary point. It could be that system $A$ reaches its lowest energy state at a finite $T_A$ while $B$ can go to lower temperatures. This would mean that $B$

A natural question, based on the previous definitions, is: to what extent can thermodynamic processes change the energy and the entropy of each system? Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if its equation of state can be sufficiently approximated by $S = \beta U = \frac{U}{k_B T}$. Internal energy is the only state variable and the only way the entropy of the system can change. We call \textbf{heat} the energy exchanged by a thermal reservoir and we note $Q = - \Delta U$ as the energy lost by the reservoir during a thermodynamic process.

>>>>

In a nutshell, we characterize processes in general terms, as the description of all possible ways that the system can evolve in time, its possible evolutions, under a particular set of circumstances, as one can see in figures \ref{fig_single_evolution} through \ref{fig_with_equilibria}. Counting the possible evolutions plays a fundamental role, in much the same way that counting states plays a role in statistical mechanics. In a deterministic and reversible process, there is a one-to-one correspondence between initial and final states: the evolutions cannot merge or diverge. If the process is deterministic but not reversible, however, the same final state will be reached from different initial states, and the evolutions will merge. If the deterministic process reaches an equilibrium, there will be no more merging. We define process entropy to be the logarithm of the count of evolutions per state, which we find to have the basic properties we associate with entropy: it is linear under system composition of independent systems, can never decrease under a deterministic process, increases if the process is non-reversible and is maximized at equilibrium. This gives us a conceptually crisp physically meaningful concept that is tied to irreversibility and equilibria in a very natural way.

@@

The paper is organized in three sections. In the first we will present a simple framework to characterize processes in general terms. In the following section we will add the definitions and assumptions that are specific to thermodynamic processes and rederive the basic laws. In the last section we see how the concepts presented relate to the different branches of mechanics.

\section{States, processes and evolutions}\label{sec_process_entropy}

Our first task is to characterize processes in the most general terms. We assume we have a system and we want to study its behavior in a given set of circumstances. If the state space $X$ of the system represents all possible configurations the system can be found in, ultimately we want to characterize how the state changes in time. As depicted in figure \ref{fig_single_evolution}, we have a copy of the state space at each moment in time, and we call an \textbf{evolution} a trajectory $\lambda : T \to X$ that for each moment in time returns the state of the system.

@@

Since not all evolutions will be allowed under all conditions, we characterize a \textbf{process} as the set of possible evolutions $E$, as we can see in figure \ref{fig_process}. Namely, a process tells us what can and cannot happen to a system in a given set of circumstances.\footnote{For example, a damped harmonic oscillator will allow trajectories whose oscillations degrade over time in a specific fashion.} We also assume we have a measure $\mu$ that allows us to determine the size of a set of evolutions, to ``count'' evolutions.\footnote{The full mathematical characterization is more complex: it will be a preorder that induces a family of measures, as to properly address sets of infinite relative sizes. However, these details do not change the overall intuition and would only distract from the key insights.} Note that a state $x_t$ at a particular time identifies a set of compatible evolutions $A(x_t) = \{ \lambda \in E \, | \, \lambda(t) = x_t \}$, those that pass through that state at that time. Using $\mu$ we can give a size to that set, which we can write as $\mu(A(x_t))$ or $\mu(x_t)$ for short.

@@

Note that the probability of transitioning from one state to another during a process is linked to the possible evolutions and their count. In fact, calculating the probability of finding the state $x_{t+\Delta t}$ at $t+\Delta t$ given $x_t$ at $t$ means determining the fraction of evolutions that pass through $x_t$ that also pass through $x_{t+\Delta t}$. That is:
\begin{equation}
	P(x_{t+\Delta t} | x_t) = \frac{\mu(A(x_t) \cap A(x_{t+\Delta t}))}{\mu(A(x_t))}.
\end{equation}
Conceptually, the ability to count evolutions is equivalent to knowing all possible transition probabilities.\footnote{Technically, a single probability measure is not enough to define all these transitions at all scales, but the overall intuition remains.}

This very minimal characterization is enough for some basic definitions. We say a process is \textbf{deterministic} if the state at one time is enough to determine the state at all future times. We can see an example of such a process in figure \ref{fig_determinism}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all future states: evolutions can merge but cannot diverge. The count of evolutions per state over time can never decrease: $
\mu(\lambda(t)) \leq \mu(\lambda(t + \Delta t))$.

Deterministic processes can be characterized by a law of evolution. That is, we can write $\lambda(t+\Delta t) = f(\lambda(t), t)$ given a suitable function $f$ which may, in general, be time dependent. We can do this because, if we fix $x_t$, $P(x_{t+\Delta t} | x_t) = 1$ for one and only one future state and the law of evolution identifies that state. Insofar we are interested in making predictions, we are implicitly restricting ourselves to deterministic processes.

@@

Conversely, we say a process is \textbf{reversible} if the state at one time is enough to reconstruct the state at all past times.\footnote{Note that in some literature the ability to reconstruct the past is called retrodictability and does not coincide with thermodynamic reversibility (the ability to undo a change). This distinction appears in the context of continuous quantities and it is one of those technical considerations we will address in a more technical work. Our definition does recover thermodynamic reversibility, without having to assume the existence of a reverse process or a symmetry under time-reversal.} We can see an example of such a process in figure \ref{fig_reversibility}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all past states: evolutions can diverge but cannot merge. The count of evolutions per state over time can never increase: $
\mu(\lambda(t)) \geq \mu(\lambda(t + \Delta t))$.

@@

A process that is both deterministic and reversible, like the one in figure \ref{fig_detrev}, will never contain evolutions that split or merge. The count of evolutions per state over time will be a constant: $\mu(\lambda(t)) = \mu(\lambda(t + \Delta t))$.

@@

To finish, consider a \textbf{deterministic processes with equilibria}, like the one in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{This characterization of equilibrium, though amply used in thermodynamic literature, is simplistic and technically does not quite work. However, it will suffice for the purpose of this paper as a full treatment would again distract from the main points.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

@@

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(\lambda(t))$, that can never decrease in a deterministic process, stays the same if the process is also reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we have two systems, with their respective state spaces $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\lambda(t) = \{\lambda_x(t), \lambda_y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in \ref{fig_independence}, for each possible evolution $\lambda_x(t)$ of the first system and $\lambda_y(t)$ of the second system, there will be a possible evolution $\{\lambda_x(t), \lambda_y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two systems are \textbf{independent} at a given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(\lambda(t)) = \mu_X (\lambda_x(t)) \mu_Y (\lambda_y(t))$.

@@

We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under a deterministic process with equilibria. This means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems.\footnote{As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$. We leave these details to future works.}

Under very minimal conditions, we have found a quantity that has all the right properties for entropy and that is linked with determinism and irreversibility in a plain, intuitive and physically meaningful way. There is one remaining issue: the process entropy will depend in general on the process, not just on the state. That is, the same description of the same system may be associated with more or fewer evolutions depending on the conditions. Entropy in physics, however, is associated to a state. Under what conditions can we associate a unique process entropy to a state?

@@

For composite systems there are additional considerations. Different processes may put different degrees of correlation between subsystems, therefore the same equilibria for the subsystems may correspond to different process entropy values for the composite. But if the subsystems are independent, the composite will correspond to a unique entropy value: the product.

We can therefore assign a unique process entropy to states of equilibria and to composite states of independent equilibria. Under these conditions we define the \textbf{state entropy} $S : X \to \mathbb{R}$ which corresponds to the process entropy that the system reaches at equilibrium. Note that both equilibrium and independence lead to entropy maximization, so the state entropy can be thought as the maximal process entropy for the system. In all other cases, the process entropy will be lower meaning that that description for that system in those conditions gives us more information because of possible correlations within the system or with other systems.\footnote{This recovers the intuition of entropy as ``lack of information'' and that entropy maximization represents the case where knowledge of the system is fully represented by the given constraints. However, in our case this is not a subjective matter, rather an objective description of what the process may or may not allow.}

This link between state entropy and independent equilibria opens a series of questions on the fundamental link between the two which go beyond the scope of this paper. However, we want at least give a partial answer to one: why is assigning a unique process entropy important? One answer is process combination. Suppose that we are studying a system and we change the circumstances surrounding it: it first evolves according to process $A$ and at $t_0$ it switches to process $B$. If $\lambda(t)$ is an evolution of the combined process $AB$, then for $t \leq t_0$ it must also be an evolution for $A$ and for $t \geq t_0$ it must also be an evolution for $B$. This means that at the junction state $x_{t_0} = \lambda(t_0)$ the count of evolution has to match: $\mu_{AB}(x_{t_0})=\mu_A(x_{t_0})=\mu_B(x_{t_0})$. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state, independent of the process. As we'll see in section \ref{sec_mechanics}, these type of consideration are already embedded in the mathematical structure of the state spaces of classical mechanics, quantum mechanics, thermodynamics and statistical mechanics.

\section{Thermodynamics}\label{sec_thermodynamics}

The above discussion sets the stage for our derivation of thermodynamics. We saw how state entropy is the maximized process entropy, a quantity that can never decrease under deterministic processes. Our task now is to extend those concepts in a way that recovers thermodynamics.

We can motivate this extension by considering process composition of composite systems. The idea is that we want to describe the situation in which a composite\footnote{Mathematically, systems under composition form a Boolean algebra. The join operation $X \cupdot Y$ returns the composite; the meet operation $X \capdot Y$ returns the smallest system contained by both; the lattice ordering $X \subseteq Y$ tells us whether $X$ is a part of $Y$ and so on.} system $X \cupdot Y$ is subject to process $A$ until time $t_0$, after which each subsystem evolves independently: $X$ according to process $B$ and $Y$ according to process $C$. At $t_0$, then, we not only have to reconcile the count of evolutions for the composite, but for each subsystem as well. Therefore $X$ and $Y$ must be independent and in equilibrium at time $t_0$. In our framework, thermodynamic equilibrium handles precisely this situation. The remaining ingredient is the notion of extensive quantities, which again serves to characterize part-whole relationships.

We define a \textbf{thermodynamic process} a particular type of deterministic process with equilibria that satisfies two additional requirements:
\begin{itemize}
	\item all parts of the systems are in equilibria and are independent from each other
	\item there is a non-trivial dependence between the initial state of one part and the final state of another (i.e. there is a way to change the initial state of one part such that the final state of another is effectively changed).
\end{itemize}
The first requirement tells us that process entropy is maximized at all levels of the system. The second requirements serves to rule out the case where the systems are isolated from each other. We say a system is in \textbf{thermodynamic equilibrium} if it is part of the output of a thermodynamic process. We say two systems are in thermodynamic equilibrium with each other if they are parts of the same system in thermodynamic equilibrium.\footnote{Mathematically, the set of all systems in the same thermodynamic equilibrium is a principal ideal $X\!\! \downarrow$.}

To get a feel for the definition, let us rederive the fact that thermodynamic equilibrium is transitive, also known as the zeroth law. Suppose system $X$ is in thermodynamic equilibrium with system $Y$ and system $Y$ is in thermodynamic equilibrium with system $Z$. Then $X$ and $Y$ are parts of the output of the same thermodynamic process and so are $Y$ and $Z$. This means that $X$ and $Z$ are parts of the output of the same thermodynamic process, and therefore are in thermodynamic equilibrium.

To characterize part-whole relationships between states, we focus on particular types of systems, which we call \textbf{thermodynamic systems}, that satisfy these additional requirements: 
\begin{itemize}
	\item the state space is formed by states of equilibria
	\item states are identified by a set of extensive quantities (i.e. additive under system composition)
	\item one such quantity, called internal energy, is conserved under all thermodynamic processes.\footnote{It may be possible that some or all of these properties can be derived from the composability of thermodynamic equilibria. Nevertheless, we consider them additional assumptions for this work.
	}
\end{itemize}
We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities.

The next step is to characterize to what extent can thermodynamic processes change the energy and the entropy of each system. Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if the internal energy $U_R$ is the only state variable. In this case, the equation of state is simply $S_R = S_R(U_R)$ and $\Delta S_R = \int \beta_R dU_R$.\footnote{The existence of unique equilibria between reservoirs forces $S_R(U_R)$ to be strictly monotonic.} We call \textbf{heat} $Q = - \Delta U_R$ the energy lost by the reservoir during a thermodynamic process.

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S_M(U_M, x^i_M) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU_M &= k_B T_M dS + X_{Mi} dx^i_M = X_{Mi} dx^i_M.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U_M$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S_M}{\partial U_M} = 0 = \beta = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

With these definitions, we can study how changes of energy and entropy are related to each other. Consider a system composed of a generic system $X$, a thermal reservoir $R$ and a mechanical system $M$.

During a thermodynamic process, because the total energy of the system must be conserved, we have:
\begin{equation}
\begin{aligned}
&\Delta U_X + \Delta U_R + \Delta U_M = 0 \\
&\Delta U_X - Q + W = 0 \\
&\Delta U_X = Q - W.
\end{aligned}
\end{equation}
In this setting, work represents the energy exchanged at constant entropy while heat represents exchanged energy associated with entropy change. This recovers the first law of thermodynamics.

We now look at the change in entropy, which during a thermodynamic process can never decrease given that thermodynamic processes are deterministic. Keeping in mind $M$ is mechanical, we have:
\begin{equation}
\begin{aligned}
&\Delta S_X + \Delta S_R + \Delta S_M \geq 0 \\
&\Delta S_X \geq - \Delta S_R \\
&\Delta S_X \geq \int \beta_R (- dU_R) \\
&\Delta S_X \geq \int \beta_R d Q \\
&k_B \Delta S_X \geq \int \frac{d Q}{T_R}.
\end{aligned}
\end{equation}
which recovers the second law of thermodynamics. This is just a more specialized version of \eqref{ov_entropy_increases} which is valid for any deterministic process.

The third law has many equivalent formulations. [CITE] One states that the entropy of any system has a non-negative lower bound which is reached by the lowest energy state. If the ground state is non-degenerate, such as a perfect crystal, its entropy is equal to zero. With some work, we can recover this form.
